---
title: "Lung Cancer"
author: "Hasan Jamil"
date: "2023-04-10"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("G:\\My Drive\\Projects\\smokingandcaner")
```

## Loading the essential libraries

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(plotly)
library(forecast)
library(prophet)
```


## Loading the datasets
```{r, message=FALSE, warning=FALSE}
# Creating a vector of file pathes 
file_pathes <- list.files(path = "G:\\My Drive\\Projects\\smokingandcaner\\datasets\\iarc", pattern = "*.csv", full.names = TRUE)
# Looping and importeing the datasets
for (i in 1:length(file_pathes)){
  assign(paste0("dataset", i), read_csv(file_pathes[i]))
}

# List all the datasets in the current working directory with a similar prefix
datasets <- ls(pattern="^dataset")

# Use lapply and rbind to combine the datasets into a single dataset
combined_data <- do.call(rbind, lapply(datasets, get))
combined_data <- combined_data %>% clean_names()
```


## Checking the data

```{r pressure, echo=FALSE}
table(combined_data$country_label)
```



## Plotting 

### Male Mortality

```{r}
male_data <- combined_data %>% filter(sex == 1)
male_plot <- ggplot(male_data) + geom_smooth(aes(year, asr_world, color=country_label)) 
ggplotly(male_plot)
```

### Female Mortality

```{r}
female_data <- combined_data %>% filter(sex == 2)
female_plot <- ggplot(female_data) + geom_smooth(aes(year, asr_world, color=country_label)) 
ggplotly(female_plot)
```

# The Forecasting
The methodology I used for forecasting involved several steps. Initially, I decided to use the Prophet package instead of ARIMA since the data was non-stationary, and Prophet has the capability to handle such data.

Next, I focused on forecasting the smoking prevalence for males in Japan using the Prophet package. I started by selecting the relevant data columns and renaming them to match the required input format for the package. Then, I converted the data to the time series format using the as.Date function. I used the resulting time series as the input to the Prophet model, and used the make_future_dataframe function to generate the future data points.

For the second step, I utilized the forecasted smoking prevalence data and the historical data of age-standardized mortality to forecast the age-standardized mortality in Japan. I incorporated the forecasted smoking prevalence data as an additional regressor in the Prophet model, as it was expected to have a significant impact on the mortality rate.

Overall, the methodology I followed involved using the Prophet package to model the non-stationary time series data, and incorporating additional regressors to improve the forecasting accuracy.
### creating the time-series for japan
  
  ```{r}
mortality <- read.csv("datasets/iarc/japan/mort.csv")
  ```


```{r}
mortality <- mortality %>% 
  clean_names() %>%
  select(year, asr_world, sex) %>%
  mutate(sex = factor(sex, labels = c("male", "female")))
```

```{r}
# plotting the asr_world per year for each sex in japan
ggplot(mortality) +
  geom_line(aes(y = asr_world, x = year, colour = sex)) +
  labs(title = "Japan Mortality Rate", x = "Year", y = "ASR Mortality")
```

```{r}
smoking <- read.csv("datasets/iarc/japan/smokingpre_updated.csv") %>% 
  rename(year = Year) %>% 
  select(year, sex, smoker) %>%
  mutate(sex = factor(sex, labels = c("male", "female")))
```

```{r}
# plotting the number of smoker per sex for each year
ggplot(smoking) +
  geom_line(aes(y = smoker, x = year, colour = sex)) +
  labs(title = "Smoking in Japan", x = "Year", y = "Prevelance of Smoking")
```


```{r}
# merging the two datasets
total_epi <- mortality %>% 
    full_join(smoking, by=c("year", "sex"))
```

```{r}
#
ggplot(total_epi) +
  geom_line(aes(y = asr_world, x = year, colour = sex)) +
    geom_line(aes(y = smoker, x = year, colour = sex)) +
  labs(title = "Japan Mortality Rate and Smoking", x = "Year")
```

```{r}
total_epi <- total_epi %>% filter(sex == "male")
```

## first forecasting the smoking prevlance for males in japan
```{r}
total_epi <- na.omit(total_epi)
df <- total_epi %>%
  select(year, smoker) %>%
  rename(ds = year, y = smoker) %>%
  mutate(ds = as.Date(paste0(ds, "-01-01"), format = "%Y-%m-%d"))

m <- prophet(df)

future <- make_future_dataframe(m, periods = 20, freq = "year")

forecast <- predict(m, future)

p <- plot(m, forecast)
p <- p + geom_vline(xintercept = forecast$ds[forecast$yhat < 0][1], linetype = "dashed", color = "red")
p <- p + labs(title = "Forecasting Smoking Prevelance in Japan", x = "Year", y = "Smoking Prevelance")
p <- p + annotate(geom = "text", x = forecast$ds[forecast$yhat < 0][1], y = 0, label = as.character(year(forecast$ds[forecast$yhat < 0][1])))
ggplotly(p)
```

## Forecasting the Lung Cancer Mortality

```{r}
total_epi_prophet <- total_epi %>%
  select(year, asr_world, smoker) %>%
  rename(ds = year, y = asr_world, regressor = smoker)
total_epi_prophet$ds <- as.Date(paste0(total_epi_prophet$ds, "-01-01"), format = "%Y-%m-%d")
```

```{r}
total_epi_prophet <- na.omit(total_epi_prophet)
```

```{r}  
m <- prophet()
m <- add_regressor(m, 'regressor')
```

```{r}
m <- fit.prophet(m, total_epi_prophet)
```

```{r}
future <- make_future_dataframe(m, periods = 20, freq = "year")
```


```{r}
total_epi <- na.omit(total_epi)
```

```{r}
future$regressor <- forecast$yhat
```

```{r}
forecast_mort <- predict(m, future)

A <- plot(m, forecast_mort)
A <- A + geom_vline(xintercept = forecast_mort$ds[forecast_mort$yhat < 0][1], linetype = "dashed", color = "red")
A <- A + labs(title = "Forecasting Lung Cancer Mortality in Japan", x = "Year", y = "Lung Cancer Mortality")
A <- A + annotate(geom = "text", x = forecast_mort$ds[forecast_mort$yhat < 0][1], y = 0, label = as.character(year(forecast_mort$ds[forecast_mort$yhat < 0][1])))
ggplotly(A)
```

```{r}

library(cowplot)

# combine the two plots horizontally
plot_combined <- plot_grid(p, A, ncol = 2)

# add title and axis labels
plot_combined <- plot_combined + ggtitle("Smoking Prevalence and Lung Cancer Mortality in Japan") +
  xlab("Year") + ylab("Rate")

# display the combined plot
plot_combined
```